---
title: "Vignobles MBV Quebec"
author: "Adriel M Sierra"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: cayman
    toc: true
    toc_depth: 3
---

# **Grapevine Endosphytes & Grape trunk diseases (GTD)**

This reproducible quarto document contains the complete statistical analysis, machine learning models, and ecological assembly null modeling workflows for evaluating grapevine trunk disease (GTD) effects on endophyte tissue dynamics and microbiote dysbiosis across 50 cold-climate vineyards in Québec, Canada.

By integrating high-throughput amplicons sequencing (16S rRNA and ITS2) with qPCR detection of major fungal wood pathogens (*Eutypa lata* and *Diplodia seriata*), this project explores how tissue compartmentalization, management practices, and host microbiota composition are related to the emergence/occurence of grapevine trunk disease (GTDs) dieback.

### Library

R-packages use for the analyses

```{r library, echo=TRUE, message=FALSE,  message=FALSE, warning=FALSE}
#Library
library(microeco)
library(tidyverse)
#remotes::install_github("ChiLiubio/mecodev")
library(mecodev)
library(stringr)
library(ggpubr)
library(vegan)
library(rstatix)
#packages for building a map
# Install if you haven't already: install.packages(c("ggplot2", "raster", "geodata", "sf"))
library(ggplot2)
library(raster)
library(sf) 
library(geodata)
#devtools::install_github("microsud/dysbiosisR")
library(dysbiosisR)

# packages used to build the quarto document
library(DT)
```

```{r wd, include=FALSE,  message=FALSE, warning=FALSE}
setwd("/Users/adrielsierra/Documents/PostDoc/INRS/Vignoable Project/Github/")
```

### Datasets

Complete and processed datasets are available as a microtable object.

```{r data,  message=FALSE, warning=FALSE}
# bacteria endophyte
MBV16S_mt = readRDS("01_MBV16S_microtable_jan26.rds")
# fungal endophyte
MBVITS_mt = readRDS("01_MBVITS_microtable_jan26.rds")
```

Microtables with alpha- and beta-diversity estimates included

```{r estimates,  message=FALSE, warning=FALSE}
MBV16S_mt_est <- readRDS("02_MBV16S_microtable_estimates_jan26.rds")
MBVITS_mt_est <- readRDS("02_MBVITS_microtable_estimates_jan26.rds")
MBV16S_alpha = readRDS("02_MBVBacterial_alphadiv.rds")
MBVITS_alpha = readRDS("02_MBVFungal_alphadiv.rds")
```

Dysbiosis score

```{r dysbiosisdata,  message=FALSE, warning=FALSE}
#datasets
dysbiosis_results_bacteria <- readRDS("03_DysbiosisBacteriaResults.rds")
dysbiosis_results_fungi <- readRDS("03_DysbiosisFungiResults.rds")
```

### Sampling design

A table include all the metadata associated for each sample.

Samples for Bacteria 16S amplicon-sequencing.

```{r sample16S,  message=FALSE, warning=FALSE}
DT::datatable(MBV16S_mt$sample_table)
```

Samples for Fungal ITS2 amplicon-sequencing

```{r sampleITS,  message=FALSE, warning=FALSE}
DT::datatable(MBV16S_mt$sample_table)
```

#### Map of vineyards analyzed across Québec

```{r}
# Map of vignobles

# Canada-wide boundary data (Level 1 = provinces/territories)
geodata_path("~/geodata")
canada_sp <- gadm(country = "CAN", level = 2)

# Filter for Quebec (assuming 'NAME_1' is the province name column)
quebec_sp <- subset(canada_sp, canada_sp$NAME_1 == "Québec")

# Convert to sf object for ggplot (optional but recommended)
quebec_sf <- st_as_sf(quebec_sp)
MBV16S_sampledata.envi <- MBV16S_mt$sample_table

MBV16S_sampledata.envi$Infected <- as.factor(
  str_replace_all(
    MBV16S_sampledata.envi$Pathologie,
    c(
      "Diplodia.seriata" = "Infected",
      "Eutypa.lata" = "Infected",
      "Eutypa.lata_Diplodia.seriata" = "Infected"
    )
  )
)
#levels(MBV16S_sampledata.envi$Infected) <- c("Healthy", "Infected", "Infected")

# plot map of all vineyards classified by symptomatic infections
ggplot() +
  geom_sf(data = quebec_sf, fill = "lightyellow", color = "black") +
  coord_sf(xlim = c(-80, -70), ylim = c(44.8, 47.5), expand = FALSE)+
  geom_point(data = MBV16S_sampledata.envi,                   # Specify the points data
             aes(x = Longitude.GPS, y = Latitude.GPS, shape = Infected, alpha = Infected, color = Pathologie), # Map aesthetics for points
             size = 3)+
  geom_jitter()+
  scale_color_manual(values = c("#1B9E77",  "#D95F02",  "#7570B3",  "NA"))+
  scale_alpha_manual(values = c("Healthy" = 0.2, "Infected" = 0.60))+
  labs(title = "Map of Québec") +
  theme_minimal()
```

### Taxonomic tables

#### Bacterial endophyte taxonomy

```{r btax,  message=FALSE, warning=FALSE}
bgenus_freq = sort(table((MBV16S_mt$tax_table)[, "Genus"], exclude = NULL), decreasing = TRUE)

top_bgenera <- names(bgenus_freq)[1:20]

tax_df <- as.data.frame(MBV16S_mt$tax_table)

tax_top_bgenera <- tax_df[tax_df$Genus %in% top_bgenera, ]

#collapse by one row per genus
tax_top_bgenera_unique <- tax_top_bgenera |>
  dplyr::select(Phylum, Class, Order, Family, Genus) |>
  dplyr::distinct()

#final table with frequencies
bacteria_final_table <- tax_top_bgenera_unique |>
  dplyr::left_join(
    data.frame(
      Genus = names(bgenus_freq),
      Frequency = as.numeric(bgenus_freq)
    ),
    by = "Genus"
  ) |>
  dplyr::arrange(desc(Frequency))

DT::datatable(bacteria_final_table)

```

#### Fungal endophyte taxonomy

```{r ftax}
fgenus_freq = sort(table((MBVITS_mt$tax_table)[, "Genus"], exclude = NULL), decreasing = TRUE)

top_fgenera <- names(fgenus_freq)[1:20]

ftax_df <- as.data.frame(MBVITS_mt$tax_table)

tax_top_fgenera <- ftax_df[ftax_df$Genus %in% top_fgenera, ]

#collapse by one row per genus
tax_top_fgenera_unique <- tax_top_fgenera |>
  dplyr::select(Phylum, Class, Order, Family, Genus) |>
  dplyr::distinct()

#final table with frequencies
fungal_final_table <- tax_top_fgenera_unique |>
  dplyr::left_join(
    data.frame(
      Genus = names(fgenus_freq),
      Frequency = as.numeric(fgenus_freq)
    ),
    by = "Genus"
  ) |>
  dplyr::arrange(desc(Frequency))

DT::datatable(fungal_final_table)

```

### Data exploration

#### Abundance differences endophyte across vineyards

-   Bacterial endophyte

```{r bactabun,  message=FALSE, warning=FALSE, message=FALSE, results = 'hide'}

# Variables plus important
#Annee.de.plantation	Mode.de.conduite	Periode.de.taille	Tissu
# Relative abundance
MBV16S_abunf <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Family", ntaxa = 6)
MBV16S_abung <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Genus")

# line plot by year planted
A_MBV16S_abunf <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Family", ntaxa = 5, groupmean = "Annee.de.plantation")
A_MBV16S_abunf$data_abund <- A_MBV16S_abunf$data_abund %>%
  dplyr::filter(!is.na(Sample))
A_MBV16S_abun <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Annee.de.plantation")
A_MBV16S_abun$data_abund <- A_MBV16S_abun$data_abund %>%
  dplyr::filter(Sample != "NA",
                Taxonomy != "NA")
# vineyards sample tissue
B_MBV16S_abunf <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Family", ntaxa = 5, groupmean = "Tissu")
B_MBV16S_abun <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Tissu")
B_MBV16S_abun$data_abund <- B_MBV16S_abun$data_abund %>%
  dplyr::filter(Taxonomy != "NA")
# vineyards training system
C_MBV16S_abunf <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Family", ntaxa = 5, groupmean = "Mode.de.conduite")
C_MBV16S_abun <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Mode.de.conduite")
C_MBV16S_abun$data_abund <- C_MBV16S_abun$data_abund %>%
  dplyr::filter(Taxonomy != "NA")
# vineyards pruning season
D_MBV16S_abunf <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Family", ntaxa = 5, groupmean = "Periode.de.taille")
D_MBV16S_abunf$data_abund <- D_MBV16S_abunf$data_abund %>%
  dplyr::filter(Sample != "")
D_MBV16S_abun <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Periode.de.taille")
D_MBV16S_abun$data_abund <- D_MBV16S_abun$data_abund %>%
  dplyr::filter(Sample != "",
                Taxonomy != "NA")

# Kruskal Wallis test genus tissue
kw_results <- MBV16S_abunf$data_abund %>%   
  group_by(Taxonomy) %>%
  kruskal_test(Abundance ~ Tissu)
#correct multiple testing
kw_results <- kw_results %>%
  mutate(p_adj = p.adjust(p, method = "BH"))
#add significance labels
kw_results <- kw_results %>%
  mutate(p.signif = case_when(
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01  ~ "**",
    p_adj <= 0.05  ~ "*",
    TRUE           ~ "ns"
  ))

#
#posthoc <- MBV16S_abunf$data_abund %>%   
#  group_by(Taxonomy) %>%
#  pairwise_wilcox_test(
#    Abundance ~ Tissu,
#    p.adjust.method = "BH", na.rm = TRUE)

# line plots
ggarrange(A_MBV16S_abunf$plot_line()+
  theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  theme(text = element_text(size = 14)), C_MBV16S_abunf$plot_line()+
    theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
    theme(text = element_text(size = 14)), B_MBV16S_abunf$plot_line()+
    theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
    theme(text = element_text(size = 14)), D_MBV16S_abunf$plot_line()+
    theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
    theme(text = element_text(size = 14)), ncol= 2, nrow = 2, common.legend = TRUE, legend = "right")

ggarrange(A_MBV16S_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
            theme(text = element_text(size = 14)), C_MBV16S_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), B_MBV16S_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), D_MBV16S_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), ncol= 2, nrow = 2, common.legend = TRUE, legend = "right")

# Infected plants

Pathos_MBV16S_abunf <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Family", ntaxa = 5, groupmean = "Pathologie")
Pathos_MBV16S_abunf$data_abund <- Pathos_MBV16S_abunf$data_abund %>%
  dplyr::filter(Sample != "Eutypa.lata_Diplodia.seriata",
                Taxonomy != "NA")

Pathos_MBV16S_abung <- trans_abund$new(dataset = MBV16S_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Pathologie")
Pathos_MBV16S_abung$data_abund <- Pathos_MBV16S_abung$data_abund %>%
  dplyr::filter(Sample != "Eutypa.lata_Diplodia.seriata",
                Taxonomy != "NA")

ggarrange(Pathos_MBV16S_abunf$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 0, vjust = 0.5))+
            theme(text = element_text(size = 14)), Pathos_MBV16S_abung$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 0, vjust = 0.5))+
            theme(text = element_text(size = 14)), ncol= 1, nrow = 2)

```

-   Fungal endophyte

```{r vineabun,  message=FALSE, warning=FALSE, message=FALSE, results = 'hide'}
# Sum sample coverage
MBVITS_mt$sample_table$coverage = MBVITS_mt$sample_sums()

# Relative abundance
MBVITS_abunf <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Family")
MBVITS_abung <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Genus")

# line plot by year planted
# vineyard year of plantation
A_MBVITS_abunf <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Family", ntaxa = 5, groupmean = "Annee.de.plantation")
A_MBVITS_abunf$data_abund <- A_MBVITS_abunf$data_abund %>%
  dplyr::filter(!is.na(Sample),
                Taxonomy != "NA")
A_MBVITS_abun <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Annee.de.plantation")
A_MBVITS_abun$data_abund <- A_MBVITS_abun$data_abund %>%
  dplyr::filter(Sample != "NA",
                Taxonomy != "NA")
# Plant tissue
B_MBVITS_abunf <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Family", ntaxa = 5, groupmean = "Tissu")
B_MBVITS_abun <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Tissu")
B_MBVITS_abun$data_abund <- B_MBVITS_abun$data_abund %>%
  dplyr::filter(Taxonomy != "NA")
# vineyard training system
C_MBVITS_abunf <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Family", ntaxa = 5, groupmean = "Mode.de.conduite")
C_MBVITS_abun <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Mode.de.conduite")
C_MBVITS_abun$data_abund <- C_MBVITS_abun$data_abund %>%
  dplyr::filter(Taxonomy != "NA")
# vineyard pruning season
D_MBVITS_abunf <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Family", ntaxa = 5, groupmean = "Periode.de.taille")
D_MBVITS_abunf$data_abund <- D_MBVITS_abunf$data_abund %>%
  dplyr::filter(Sample != "")
D_MBVITS_abun <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Periode.de.taille")
D_MBVITS_abun$data_abund <- D_MBVITS_abun$data_abund %>%
  dplyr::filter(Sample != "",
                Taxonomy != "NA")



# Kruskal Wallis test genus tissue
kw_results <- MBVITS_abunf$data_abund %>%   
  group_by(Taxonomy) %>%
  kruskal_test(Abundance ~ Sample)
#correct multiple testing
kw_results <- kw_results %>%
  mutate(p_adj = p.adjust(p, method = "BH"))
#add significance labels
kw_results <- kw_results %>%
  mutate(p.signif = case_when(
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01  ~ "**",
    p_adj <= 0.05  ~ "*",
    TRUE           ~ "ns"
  ))

# post hot test
#posthoc <- MBVITS_abunf$data_abund %>%   
#  group_by(Taxonomy) %>%
#  pairwise_wilcox_test(
#    Abundance ~ Sample,
#    p.adjust.method = "BH")

#check abundance distribution of significant taxa
ggarrange(A_MBVITS_abunf$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
            theme(text = element_text(size = 14)), C_MBVITS_abunf$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), B_MBVITS_abunf$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), D_MBVITS_abunf$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), ncol= 2, nrow = 2, common.legend = TRUE, legend = "right")

ggarrange(A_MBVITS_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
            theme(text = element_text(size = 14)), C_MBVITS_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), B_MBVITS_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), D_MBVITS_abun$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 15, vjust = 0.5))+
            theme(text = element_text(size = 14)), ncol= 2, nrow = 2, common.legend = TRUE, legend = "right")

# Infected plants

Pathos_MBVITS_abunf <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Family", ntaxa = 5, groupmean = "Pathologie")
Pathos_MBVITS_abunf$data_abund <- Pathos_MBVITS_abunf$data_abund %>%
  dplyr::filter(Sample != "Eutypa.lata_Diplodia.seriata",
                Taxonomy != "NA")

Pathos_MBVITS_abung <- trans_abund$new(dataset = MBVITS_mt, taxrank = "Genus", ntaxa = 6, groupmean = "Pathologie")
Pathos_MBVITS_abung$data_abund <- Pathos_MBVITS_abung$data_abund %>%
  dplyr::filter(Sample != "Eutypa.lata_Diplodia.seriata",
                Taxonomy != "NA")

ggarrange(Pathos_MBVITS_abunf$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 0, vjust = 0.5))+
            theme(text = element_text(size = 14)), Pathos_MBVITS_abung$plot_line()+
            theme_classic()+ theme(axis.text.x = element_text(angle = 0, vjust = 0.5))+
            theme(text = element_text(size = 14)), ncol= 1, nrow = 2)

```

### Alpha-diversity

General script for alpha diversity estimates

```{r alpha, eval=FALSE, message=FALSE,  message=FALSE, warning=FALSE}
# Diversity 
### Alpha diversity general 
set.seed(6565)

# general script to estimate diversity index for bacterial and fungal endophytes using Hill numbers
MBV_alpha <- trans_alpha$new(dataset = MBV_mt, group = "Tissu", by_group = "Infected", by_ID = "Echantillon")# group by tissue sampled
# transform taxa occurence to relative abundance
MBV_mt_norm <- microbiome::transform(MBV_mt$otu_table, transform = "compositional")

###alpha diversity
#observe species hill numbers
hillq0 <- subset(MBV_alpha$data_alpha, MBV_alpha$data_alpha$Measure == "Observed")

# Shannon 
shannon.hill <- exp(vegan::diversity(t(MBV_mt_norm), index = "shannon"))

# Simpson
simpson.hill <- 1/(1-(vegan::diversity(t(MBV_mt_norm), index = "simpson")))
```

The Shapiro-Wilk test and histogram visualization to check for diversity estimates are normal distributed.

-   Bacteria

```{r, results = 'hide', message=FALSE,  message=FALSE, warning=FALSE}
# Statistics of both bacteria and fungi
# check for normality with histrogram
ggarrange(gghistogram(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Observed"), "Value", y = "count", fill = "Measure",
            add = "mean"),
gghistogram(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Shannon_hill"), "Value", y = "count", fill = "Measure",
            add = "mean"),
gghistogram(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Simpson_hill"), "Value", y = "count", fill = "Measure",
            add = "mean"), nrow = 1, ncol = 3)

#Bacteria
#data.frame(shapiro.test(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Observed")[,3])[1:4],
#shapiro.test(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Shannon_hill")[,3])[1:4],
#shapiro.test(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Simpson_hill")[,3])[1:4])
```

-   Fungi

```{r, results = 'hide', message=FALSE,  message=FALSE, warning=FALSE }
### Fungi
ggarrange(gghistogram(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Observed"), "Value", y = "count", fill = "Measure",
            add = "mean"),
gghistogram(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Shannon_hill"), "Value", y = "count", fill = "Measure",
            add = "mean"),
gghistogram(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Simpson_hill"), "Value", y = "count", fill = "Measure",
            add = "mean"), nrow = 1, ncol = 3)

#data.frame(shapiro.test(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Observed")[,3])[1:4],
#                     shapiro.test(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Shannon_hill")[,3])[1:4],
#                     shapiro.test(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Simpson_hill")[,3])[1:4])
```

#### Alpha-diversity statistical analyses

```{r, results = 'hide', message=FALSE,  message=FALSE, warning=FALSE}
# Alpha diversity summary statistics
########
# estimate mean and standard deviation for groups
data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
  return(data_sum)
}   

#
Bacterialmeandiv = data.frame(dfrich.obs.b <- data_summary(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Observed"), varname="Value", groupnames=c("Annee.echantillonnage.fact", "Tissu", "Infected")),
                           dfrich.sha.b <- data_summary(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Shannon_hill"), varname="Value", groupnames=c("Annee.echantillonnage.fact", "Tissu", "Infected")),
                           dfrich.simp.b <- data_summary(subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Simpson_hill"), varname="Value", groupnames=c("Annee.echantillonnage.fact", "Tissu", "Infected")))

Fungalmeandiv = data.frame(dfrich.obs.plot <- data_summary(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Observed"), varname="Value", groupnames=c("Annee.echantillonnage.fact", "Tissu", "Infected")),
                     dfrich.sha.plot <- data_summary(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Shannon_hill"), varname="Value", groupnames=c("Annee.echantillonnage.fact", "Tissu", "Infected")),
                     dfrich.simp.plot <- data_summary(subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Simpson_hill"), varname="Value", groupnames=c("Annee.echantillonnage.fact", "Tissu", "Infected")))


summary(glm(Value~Tissu*Infected, family = Gamma(link = "log"), data=subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Observed")))
summary(glm(Value~Tissu*Infected, family = Gamma(link = "log"), data=subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Shannon_hill")))
summary(glm(Value~Tissu*Infected, family = Gamma(link = "log"), data=subset(MBV16S_alpha$data_alpha, MBV16S_alpha$data_alpha$Measure == "Simpson_hill")))

summary(glm(Value~Tissu*Infected, family = Gamma(link = "log"), data=subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Observed")))
summary(glm(Value~Tissu*Infected, family = Gamma(link = "log"), data=subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Shannon_hill")))
summary(glm(Value~Tissu*Infected, family = Gamma(link = "log"), data=subset(MBVITS_alpha$data_alpha, MBVITS_alpha$data_alpha$Measure == "Simpson_hill")))
```

#### Visualize alpha-diversity

```{r, results = 'hide',  message=FALSE, warning=FALSE }

#Visualize
#bacteria
library(introdataviz)
library(devtools)
#devtools::install_github("psyteachr/introdataviz")
library(purrr)
library(emmeans)
library(multcomp)


div_measures <- c("Observed", "Shannon_hill", "Simpson_hill")
levels(MBV16S_alpha$data_alpha$Tissu) = c("Ecorce.Chancre", "Courson.Sarment","Cordon")

plot_list_bacteria <- list()

for (measure in div_measures) {
    
    df_sub <- MBV16S_alpha$data_alpha %>%
      filter(Measure == measure)
  # ---- Fit GLM ----
  model <- glm(Value ~ Tissu * Infected,
               data = df_sub,
               family = Gamma(link = "log"))
  
  # ---- Estimated marginal means ----
  emm <- emmeans(model, ~ Tissu * Infected)
  
  # ---- Compact letter display ----
  cld_res <- multcomp::cld(emm, Letters = letters) %>%
    as.data.frame()
  cld_res$maxval = max(df_sub$Value)
  # ---- Plot ----
  p <- ggplot(df_sub,
              aes(x = Tissu,
                  y = Value,
                  fill = Infected)) +
    
    geom_violin(trim = FALSE) +
    
    geom_jitter(position = position_jitterdodge(
                  dodge.width = 0.8,
                  jitter.width = 0.6),
                size = 1.5) +
    
    geom_text(data = cld_res,
              aes(x = Tissu,
                  y = maxval+100,
                  label = .group,
                  group = Infected),
              position = position_dodge(0.8),
              inherit.aes = FALSE,
              size = 5) +
    labs(x = "Vine tissue",
         y = measure,
         title = measure) +
    scale_fill_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 14))
  
  plot_list_bacteria[[measure]] <- p
}

plot_list_bacteria[[1]]
plot_list_bacteria[[2]]
plot_list_bacteria[[3]]

### plot fungi
MBVITS_alpha_sub <- NULL
MBVITS_alpha_sub$data_alpha <- MBVITS_alpha$data_alpha %>%
  filter(Tissu != "Racine")

plot_list_fungi <- list()
for (measure in div_measures) {
  
  df_sub <- MBVITS_alpha_sub$data_alpha %>%
    filter(Measure == measure)
  # ---- Fit GLM ----
  model <- glm(Value ~ Tissu * Infected,
               data = df_sub,
               family = Gamma(link = "log"))
  
  # ---- Estimated marginal means ----
  emm <- emmeans(model, ~ Tissu * Infected)
  
  # ---- Compact letter display ----
  cld_res <- multcomp::cld(emm, Letters = letters) %>%
    as.data.frame()
  cld_res$maxval = max(df_sub$Value)
  # ---- Plot ----
  p <- ggplot(df_sub,
              aes(x = Tissu,
                  y = Value,
                  fill = Infected)) +
    
    geom_violin(trim = FALSE) +
    
    geom_jitter(position = position_jitterdodge(
      dodge.width = 0.8,
      jitter.width = 0.6),
      size = 1.5) +
    
    geom_text(data = cld_res,
              aes(x = Tissu,
                  y = maxval+10,
                  label = .group,
                  group = Infected),
              position = position_dodge(0.8),
              inherit.aes = FALSE,
              size = 5) +
    labs(x = "Vine tissue",
         y = measure,
         title = measure) +
    scale_fill_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 14))
  
  plot_list_fungi[[measure]] <- p
}

plot_list_fungi[[1]]
plot_list_fungi[[2]]
plot_list_fungi[[3]]
```

### Beta-diversity

```{r, results = 'hide',  message=FALSE, warning=FALSE }
#########Beta diversity 
# Betadiversity Analysis by tissue as principal group
#"Tissu"
#Bacteria
MBV16S_mt$cal_betadiv()
bact.beta <- trans_beta$new(dataset = MBV16S_mt, group = "Tissu", measure = "bray")

bact.beta$cal_ordination(method = "PCoA")
# $res_ordination is the ordination result list

#bacteria plot by tissue
bact.beta$plot_ordination(plot_color = "Tissu", plot_shape = "Tissu", plot_type = c("point", "ellipse"))+
  theme_classic() +
  theme(text = element_text(size = 14))

#Fungal
MBVITS_mt$sample_table = MBVITS_mt$sample_table %>%
  filter(Tissu != "Racine")
MBVITS_mt$tidy_dataset()

MBVITS_abun <- trans_abund$new(dataset = MBVITS_mt)
MBVITS_mt$cal_betadiv()

fung.beta <- trans_beta$new(dataset = MBVITS_mt, group = "Tissu", measure = "bray")

fung.beta$cal_ordination(method = "PCoA")
# t1$res_ordination is the ordination result list


#envfit (community-level effects) bacteria
#Ordination 
ord_scores_bact <- scores(bact.beta$res_ordination, display = "sites")

MBVenvi.data.vars<-ord_scores_bact[,c(19:20,27:38)]
MBVenvi.data.vars$dysbiosis <- dysbiosis_results_bacteria$dysbiosisOOBRF$oob.score
# Fit environmental data

fit_bact <- envfit(ord_scores_bact[,c(1:2)], MBVenvi.data.vars, permutations = 999, na.rm = TRUE)
#fit_bact_sig<-data.frame(fit_bact$vectors)

# Extract the arrow coordinates
# These are the x and y directions of the arrows
arrows_bact <- as.data.frame(scores(fit_bact, display = "vectors"))
arrows_bact$r <- fit_bact$vectors$r
arrows_bact$pvals <- fit_bact$vectors$pvals
arrows_bact$vars <- rownames(arrows_bact)
arrows_bact_sig <- arrows_bact[which(arrows_bact$pvals < 0.06),]
# Multiply arrows by a factor so they are visible 
# (sometimes they are too small/large relative to the points)
scaling_factor <- 0.8 
arrows_bact[,1:2] <- arrows_bact[,1:2] * scaling_factor

#envfit (community-level effects) fungi
#Ordination 
ord_scores_fung <- scores(fung.beta$res_ordination, display = "sites")


MBVenvi.data.vars.fung<-ord_scores_fung[,c(19:20,27:38)]

dysbiosis_results_fungi$dysbiosisOOBRF = dysbiosis_results_fungi$dysbiosisOOBRF %>%
  filter(Tissu != "Racine")
MBVenvi.data.vars.fung$dysbiosis <- dysbiosis_results_fungi$dysbiosisOOBRF$oob.score
ord_scores_fung$dysbiosisOOBRF = dysbiosis_results_fungi$dysbiosisOOBRF$oob.score
# Fit environmental datas

fit_fung <- envfit(ord_scores_fung[,c(1:2)], MBVenvi.data.vars.fung, permutations = 999, na.rm = TRUE)


# Extract the arrow coordinates
# These are the x and y directions of the arrows
arrows_fung <- as.data.frame(scores(fit_fung, display = "vectors"))
arrows_fung$r <- fit_fung$vectors$r
arrows_fung$pvals <- fit_fung$vectors$pvals
arrows_fung$vars <- rownames(arrows_fung)
arrows_fung_sig <- arrows_bact[which(arrows_fung$pvals < 0.06),]
# Multiply arrows by a factor so they are visible 
# (sometimes they are too small/large relative to the points)
scaling_factor <- 0.8 
arrows_fung[,1:2] <- arrows_fung[,1:2] * scaling_factor
# plot the PCoA result with confidence ellipse
#for bacteria and fingi
ggarrange(bact.beta$plot_ordination(plot_color = "Tissu", plot_shape = "Tissu", plot_type = c("point", "ellipse"))+
            geom_segment(data = arrows_bact_sig, 
                         aes(x = 0, y = 0, xend = PCo1, yend = PCo2), 
                         arrow = arrow(length = unit(0.2, "cm")), 
                         color = "black", inherit.aes = FALSE) +
            geom_text(data = arrows_bact_sig, 
                      aes(x = PCo1 * 1.1, y = PCo2 * 1.1, label = vars), 
                      color = "black", fontface = "bold", inherit.aes = FALSE) +
  theme_classic() +
  theme(text = element_text(size = 14)), fung.beta$plot_ordination(plot_color = "Tissu", plot_shape = "Tissu", plot_type = c("point", "ellipse"))+
    geom_segment(data = arrows_fung_sig, 
                aes(x = 0, y = 0, xend = PCo1, yend = PCo2), 
                arrow = arrow(length = unit(0.2, "cm")), 
                color = "black", inherit.aes = FALSE) +
    geom_text(data = arrows_fung_sig, 
              aes(x = PCo1 * 1.1, y = PCo2 * 1.1, label = vars), 
              color = "black", fontface = "bold", inherit.aes = FALSE) +
    theme_classic() +
  theme(text = element_text(size = 14)), ncol = 2, labels = c("Bacteria", " Fungi"), common.legend = TRUE, legend = "right")


#### Loop for generating ordination plots by tissues

# Define tissues
tissues <- c("Cordon", "Courson.Sarment", "Ecorce.Chancre")
MBV16S_mt$sample_table$dysbiosis <- dysbiosis_results_bacteria$dysbiosisOOBRF$oob.score


# Empty list to store plots for bacteria
ordination_plots_bact <- list()

for (tissue in tissues) {
  
  # Filter sample table
  sub_micro <-clone(MBV16S_mt)
  sub_micro$sample_table <- subset(sub_micro$sample_table, Tissu == tissue)
  sub_micro$cal_betadiv()
  bact.beta.sub <- trans_beta$new(dataset = sub_micro, group = "Infected", measure = "bray")
  
  bact.beta.sub$cal_ordination(method = "PCoA")
    # Optional: inspect tidy dataset
  df <- sub_micro$tidy_dataset
  print(paste(tissue, "samples:", nrow(df)))
  #enfit
  #Ordination 
  ord_scores_bact <- scores(bact.beta.sub$res_ordination, display = "sites")
  
  MBVenvi.data.vars<-ord_scores_bact[,c(19:20,27,29:38,43)]
  # 2. Fit environmental datas
  
  fit_bact <- envfit(ord_scores_bact[,c(1:2)], MBVenvi.data.vars, permutations = 999, na.rm = TRUE)
  # 3. Extract the arrow coordinates
  # These are the x and y directions of the arrows
  arrows_bact <- as.data.frame(scores(fit_bact, display = "vectors"))
  arrows_bact$r <- fit_bact$vectors$r
  arrows_bact$pvals <- fit_bact$vectors$pvals
  arrows_bact$vars <- rownames(arrows_bact)
  arrows_bact_sigt <- arrows_bact[which(arrows_bact$pvals < 0.06),]
  # 4. Optional: Multiply arrows by a factor so they are visible 
  # (sometimes they are too small/large relative to the points)
  scaling_factor <- 0.8 
  arrows_bact_sigt[,1:2] <- arrows_bact_sigt[,1:2] * scaling_factor
# Generate ordination plot
  p <- bact.beta.sub$plot_ordination(
    plot_color = "Infected",
    plot_shape = "Infected",
    plot_type = c("point", "ellipse")
  )+
    geom_segment(data = arrows_bact_sigt, 
                 aes(x = 0, y = 0, xend = PCo1, yend = PCo2), 
                 arrow = arrow(length = unit(0.2, "cm")), 
                 color = "black", inherit.aes = FALSE) +
    geom_text(data = arrows_bact_sigt, 
              aes(x = PCo1 * 1.1, y = PCo2 * 1.1, label = vars), 
              color = "black", fontface = "bold", inherit.aes = FALSE) +
    labs(title = tissue) +
    scale_color_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 16))
  
  # Store in list
  ordination_plots_bact[[tissue]] <- p
}

# display the Cordon plot
ordination_plots_bact$Cordon
ordination_plots_bact$Courson.Sarment
ordination_plots_bact$Ecorce.Chancre

# Empty list to store plots for fungi
MBVITS_mt$sample_table$dysbiosis <- dysbiosis_results_fungi$dysbiosisOOBRF$oob.score


ordination_plots_fung <- list()

for (tissue in tissues) {
  # Filter sample table
  sub_micro <-clone(MBVITS_mt)
  sub_micro$sample_table <- subset(sub_micro$sample_table, Tissu == tissue)
  sub_micro$cal_betadiv()
  fung.beta.sub <- trans_beta$new(dataset = sub_micro, group = "Infected", measure = "bray")
  
  fung.beta.sub$cal_ordination(method = "PCoA")
  # Optional: inspect tidy dataset
  df <- sub_micro$tidy_dataset
  print(paste(tissue, "samples:", nrow(df)))
  #enfit
  #Ordination 
  ord_scores_fung <- scores(fung.beta.sub$res_ordination, display = "sites")
  colnames(ord_scores_fung)
  MBVenvi.data.vars<-ord_scores_fung[,c(19:20,27,29:38,41)]
  # Fit environmental datas
  fit_fung <- envfit(ord_scores_fung[,c(1:2)], MBVenvi.data.vars, permutations = 999, na.rm = TRUE)
  # 3. Extract the arrow coordinates
  # These are the x and y directions of the arrows
  arrows_fung <- as.data.frame(scores(fit_fung, display = "vectors"))
  arrows_fung$r <- fit_fung$vectors$r
  arrows_fung$pvals <- fit_fung$vectors$pvals
  arrows_fung$vars <- rownames(arrows_fung)
  arrows_fung_sigt <- arrows_fung[which(arrows_fung$pvals < 0.06),]
  #  Multiply arrows by a factor so they are visible 
  # (sometimes they are too small/large relative to the points)
  scaling_factor <- 0.8 
  arrows_fung_sigt[,1:2] <- arrows_fung_sigt[,1:2] * scaling_factor
  # Generate ordination plot
  p <- fung.beta.sub$plot_ordination(
    plot_color = "Infected",
    plot_shape = "Infected",
    plot_type = c("point", "ellipse")
  )+
    geom_segment(data = arrows_fung_sigt, 
                 aes(x = 0, y = 0, xend = PCo1, yend = PCo2), 
                 arrow = arrow(length = unit(0.2, "cm")), 
                 color = "black", inherit.aes = FALSE) +
    geom_text(data = arrows_fung_sigt, 
              aes(x = PCo1 * 1.1, y = PCo2 * 1.1, label = vars), 
              color = "black", fontface = "bold", inherit.aes = FALSE) +
    xlim(-0.75, NA)+labs(title = tissue) +
    scale_color_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 16))
  
  # Store in list
  ordination_plots_fung[[tissue]] <- p
}

# display the Cordon plot

ordination_plots_fung$Cordon
ordination_plots_fung$Courson.Sarment
ordination_plots_fung$Ecorce.Chancre
```

#### Betadiversity statistical analyses and ordination distances within groups

```{r, results = 'hide',  message=FALSE, warning=FALSE  }

# calculate and plot sample distances within groups

# Empty list to store plots for bacteria
distance_plots_bact <- list()

for (tissue in tissues) {
  # Filter sample table
  sub_micro <-clone(MBV16S_mt)
  sub_micro$sample_table <- subset(sub_micro$sample_table, Tissu == tissue)
  sub_micro$cal_betadiv()
  bact.beta.sub <- trans_beta$new(dataset = sub_micro, group = "Infected", measure = "bray")
  bact.beta.sub$cal_group_distance(within_group = TRUE)
  bact.beta.sub$cal_group_distance_diff(method = "wilcox")
  # plot_group_order parameter can be used to adjust orders in x axis
  # Store in list
  distance_plots_bact[[tissue]] <- bact.beta.sub$plot_group_distance(add = "mean")+
    labs(title = tissue) +
    scale_color_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 14))
}  
#fungi
distance_plots_fung <- list()

for (tissue in tissues) {
  
  # Filter sample table
  sub_micro <-clone(MBVITS_mt)
  sub_micro$sample_table <- subset(sub_micro$sample_table, Tissu == tissue)
  sub_micro$cal_betadiv()
  fung.beta.sub <- trans_beta$new(dataset = sub_micro, group = "Infected", measure = "bray")
  fung.beta.sub$cal_group_distance(within_group = TRUE)
  fung.beta.sub$cal_group_distance_diff(method = "wilcox")
  # plot_group_order parameter can be used to adjust orders in x axis
  # Store in list
  distance_plots_fung[[tissue]] <- fung.beta.sub$plot_group_distance(add = "mean")+
    labs(title = tissue) +
    scale_color_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 14))
}

ggarrange(ggarrange(plotlist = distance_plots_bact, nrow = 1, ncol = 3, common.legend = TRUE), 
          ggarrange(plotlist = distance_plots_fung,nrow = 1, ncol = 3, common.legend = TRUE), 
          nrow = 2, labels = c("A", "B"))


bact.beta$cal_group_distance(within_group = FALSE)

# perform Wilcoxon Rank Sum and Signed Rank Tests
bact.beta$cal_group_distance_diff(method = "wilcox")

# calculate and plot sample distances within groups
fung.beta$cal_group_distance(within_group = TRUE)
# perform Wilcoxon Rank Sum and Signed Rank Tests
fung.beta$cal_group_distance_diff(method = "wilcox")

# Bacteria
# Permanova
bact.beta$cal_manova(manova_all = FALSE, group = "Tissu", by_group = "Infected")
## Multivariate homogeneity of groups dispersions (variances)
bact.beta$cal_betadisper()
bact.beta$res_betadisper
#Fungi
#Permanova
fung.beta$cal_manova(manova_all = FALSE, group = "Tissu", by_group = "Infected")
fung.beta$res_manova
## Multivariate homogeneity of groups dispersions (variances)
#
fung.beta$cal_betadisper()
fung.beta$res_betadisper
#PERMANOVA
# manova for all groups when manova_all = TRUE
bact.beta$cal_manova(manova_all = TRUE)
bact.beta$res_manova

fung.beta$cal_manova(manova_all = TRUE)
fung.beta$res_manova
```

### Differential abundance analyses

```{r, results = 'hide',  message=FALSE, warning=FALSE }
# Random forest feature selection
# Tissue discriminate taxa
# use Genus level for parameter taxa_level, if you want to use all taxa, change to "all"
# nresam = 1 and boots = 1 represent no bootstrapping and use all samples directly
#bacteria
bact.diff.rf <- trans_diff$new(dataset = MBV16S_mt, method = "rf", group = "Tissu", taxa_level = "Genus")
# plot the MeanDecreaseGini bar
# group_order is designed to sort the groups
g1b <- bact.diff.rf$plot_diff_bar(use_number = 1:20, group_order = NULL)+
  theme_classic() +
  theme(text = element_text(size = 16))
# plot the abundance using same taxa in g1
g2b <- bact.diff.rf$plot_diff_abund(group_order = NULL, select_taxa = bact.diff.rf$plot_diff_bar_taxa, plot_type = "barerrorbar", add_sig = FALSE, errorbar_addpoint = FALSE, errorbar_color_black = TRUE)+
  theme_classic() +
  theme(text = element_text(size = 16))
# now the y axis in g1 and g2 is same, so we can merge them
# remove g1 legend; remove g2 y axis text and ticks
g1b <- g1b + theme(legend.position = "none")
g2b <- g2b + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.border = element_blank())
pb <- g1b %>% aplot::insert_right(g2b)
pb

#fungi
fungi.diff.rf <- trans_diff$new(dataset = MBVITS_mt, method = "rf", group = "Tissu", taxa_level = "Genus")
# plot the MeanDecreaseGres_diff# plot the MeanDecreaseGini bar
# group_order is designed to sort the groups
g1f <- fungi.diff.rf$plot_diff_bar(use_number = 1:20, group_order = NULL)
# plot the abundance using same taxa in g1
g2f <- fungi.diff.rf$plot_diff_abund(group_order = NULL, select_taxa = fungi.diff.rf$plot_diff_bar_taxa, plot_type = "barerrorbar", add_sig = FALSE, errorbar_addpoint = FALSE, errorbar_color_black = TRUE)
# now the y axis in g1 and g2 is same, so we can merge them
# remove g1 legend; remove g2 y axis text and ticks
g1f <- g1f + theme(legend.position = "none")
g2f <- g2f + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.border = element_blank())
pf <- g1f %>% aplot::insert_right(g2f)
pf

bacterial_res_rf <- list()
#filtering by tissue to identify differential taxa on infected vs healthy plants
tissues <- c("Cordon", "Courson.Sarment", "Ecorce.Chancre")

for (tissue in tissues) {
  # Filter sample table
  sub_microt <-clone(MBV16S_mt)
  sub_microt$sample_table <- subset(sub_microt$sample_table, Tissu == tissue)
  sub_microt$tidy_dataset()
  bacterial_res_rf[[tissue]] <-  trans_diff$new(dataset = sub_microt, method = "rf", group = "Infected", taxa_level = "Genus", p_adjust_method = "none")
}

fungal_res_rf <- list()
#filtering by tissue to identify differential taxa on infected vs healthy plants
for (tissue in tissues) {
  
  # Filter sample table
  sub_microt <-clone(MBVITS_mt)
  sub_microt$sample_table <- subset(sub_microt$sample_table, Tissu == tissue)
  sub_microt$tidy_dataset()
  #fungal_res_rf[[tissue]] <-  trans_diff$new(dataset = sub_microt, method = "metagenomeSeq", group = "Infected", taxa_level = "Family", p_adjust_method = "none")
  fungal_res_rf[[tissue]] <-  trans_diff$new(dataset = sub_microt, method = "rf", group = "Infected", taxa_level = "Genus", p_adjust_method = "none")
}

# plot fungal features descriminating healthy and symptomatic vines in Cordon tissue
ggarrange(fungal_res_rf$Cordon$plot_diff_bar(color_values = "black")+ 
            labs(title = "Cordon") +
            #theme(legend.position = "none")+
    scale_fill_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 16)), fungal_res_rf$Cordon$plot_diff_abund(coord_flip = F, use_number = c(1:2), group_order = NULL)+ 
    scale_color_brewer(palette = "Accent") +
    theme_classic() +
    theme(text = element_text(size = 16)), ncol = 2, labels = c("A", "B"), common.legend = TRUE, legend = "right")




```

### Dysbiosis measure

**Random Forest Prediction Based Score using the dysbiosisOBB function**

```{r dysbiosis, results = 'hide',  message=FALSE, warning=FALSE}
# Dysbiosis index
# This analysis ask the question : how far does a sample fall from the references samples "healthy" microbial neighborhood?

# Dysbiosis R package
#it uses a phyloseq object so we need to transform

tax_matbact <- as.matrix(MBV16S_mt$tax_table [,c(2:7)])
MBV16S_ps = phyloseq(otu_table(MBV16S_mt$otu_table, taxa_are_rows = TRUE), tax_table(tax_matbact), sample_data(MBV16S_mt$sample_table))


tax_mat <- as.matrix(MBVITS_mt$tax_table [,c(2:7)])
MBVITS_ps = phyloseq(otu_table(MBVITS_mt$otu_table, taxa_are_rows = TRUE), tax_table(tax_mat), sample_data(MBVITS_mt$sample_table))
subset_samples(MBVITS_ps, Tissu =="Cordon")

#transform data
totalb = median(sample_sums(MBV16S_ps))
totalf = median(sample_sums(MBVITS_ps))
standfb = function(x, t=totalb) round(t * (x / sum(x)))
standff = function(x, t=totalf) round(t * (x / sum(x)))
MBV16S_ps_ra = transform_sample_counts(MBV16S_ps, standfb)
MBVITS_ps_ra = transform_sample_counts(MBVITS_ps, standff)
#sampling sizes
#table(sample_data(MBV16S_ps_ra)$Infected)
#table(sample_data(MBVITS_ps_ra)$Infected)


###
#
#Dysbiosis measures
# Compare area under the curve
# use one dataset and then save in a list at the end of the script for all indexes
# REMEMBER change the phyloseq object to select eihter bacteria or fungal dataset
##
dist.mat <- phyloseq::distance(MBV16S_ps_ra, "bray")
# get reference samples
ref.samples <- sample_names(subset_samples(MBV16S_ps_ra, 
                                           Infected == "Healthy"))
# random forest prediction based score

# data are relative abundances summed to 100 or
# near 100 val
dysbiosis.oob <- dysbiosisOBB(MBV16S_ps_ra,
                              group_col = "Infected",
                              case_label = "Infected",
                              seed_value = 1235,
                              add_tuneRF_params = list(ntreeTry=100,
                                                       stepFactor=1.5,
                                                       improve=0.01,
                                                       trace=TRUE,
                                                       dobest=FALSE),
                              ntree = 100,
                              plot_roc = TRUE)

p5 <- plotDysbiosis(df=dysbiosis.oob,
                    xvar="Infected",
                    yvar="oob.score",
                    colors=c(Infected="brown3", Healthy="steelblue"),
                    show_points = FALSE)+
  labs(x="", y="OOB Dysbiosis Score")
p5

###Gradrient plot

# define gradient cols

volcano <- c("#003f5c", "#58508d","#bc5090","#ff6361", "#ffa600")

# RF OOB  
p.obb <- plotDysbiosisGradient(df=dysbiosis.oob,
                               score="oob.score",
                               high_line = 0.5,
                               # low_line = normobiosis_thres,
                               group_var = "Infected",
                               group_colors=c("Healthy" = "steelblue", 
                                              "Infected"= "brown3"),
                               point_size = 2,
                               bg_colors = rev(volcano),
                               jitter_width = 0.1) +
  labs(y="Dysbiosis Score", subtitle = "RF OOB Probability") +
  # adjust the x and y values to fit to plot
  ggplot2::annotate("text", x = 0.7, y = 0.55,
                    label = "OOB probability\n(0.5)", color="white")

#Visualize dysbiosis scores
ggarrange(plotDysbiosisGradient(df=dysbiosis_results_bacteria$dysbiosisOOBRF,
                                score="oob.score",
                                high_line = 0.5,
                                # low_line = normobiosis_thres,
                                group_var = "Infected",
                                group_colors=c("Healthy" = "steelblue", 
                                               "Infected"= "brown3"),
                                point_size = 6,
                                bg_colors = rev(volcano),
                                jitter_width = 0.1) +
            labs(y="Dysbiosis Score", x= "Vine Individuals", title = "Bacterial Endophyte", subtitle = "RF OOB Probability") +theme_classic() +
            theme(text = element_text(size = 16))+
            # adjust the x and y values to fit to plot
            ggplot2::annotate("text", x = 0.7, y = 0.55,
                              label = "OOB probability\n(0.5)", color="white"),
plotDysbiosisGradient(df=dysbiosis_results_fungi$dysbiosisOOBRF,
                      score="oob.score",
                      high_line = 0.5,
                      # low_line = normobiosis_thres,
                      group_var = "Infected",
                      group_colors=c("Healthy" = "steelblue", 
                                     "Infected"= "brown3"),
                      point_size = 6,
                      bg_colors = rev(volcano),
                      jitter_width = 0.1) +
  labs(y="Dysbiosis Score", x= "Vine Individuals", title = "Fungal Endophyte", subtitle = "RF OOB Probability") +theme_classic() +
  theme(text = element_text(size = 16))+
  # adjust the x and y values to fit to plot
  ggplot2::annotate("text", x = 0.7, y = 0.55,
                    label = "OOB probability\n(0.5)", color="white"), common.legend = TRUE, legend = "right")


### use interactive plot to identified samples metadata on the plot
library(plotly)
#install.packages("plotly")
# Convert the ggplot object to an interactive plotly graph
ggplotly(p.obb, originalData = TRUE)

boxplot(dysbiosis.oob$oob.score~dysbiosis.oob$Tissu)

```

Mapping dysbiosis score across Québec vineyards

```{r mapdysbiosis, results = 'hide',  message=FALSE, warning=FALSE}
#Dysbiosis map
#dysbiosis_results_bacteria$dysbiosisOOBRF
bmeta_dys_inf <- subset(dysbiosis_results_bacteria$dysbiosisOOBRF, dysbiosis_results_bacteria$dysbiosisOOBRF$Infected == "Infected")

fmeta_dys_inf <- subset(dysbiosis_results_fungi$dysbiosisOOBRF, dysbiosis_results_fungi$dysbiosisOOBRF$Infected == "Infected")

ggplot() +
  ggtitle("Bacterial endophyte")+
  geom_sf(data = quebec_sf, fill = "lightblue", color = "black") +
  coord_sf(xlim = c(-80, -70), ylim = c(44.8, 47.5), expand = FALSE)+
  geom_point(data = bmeta_dys_inf,                   # Specify the points data
             aes(x = Longitude.GPS, y = Latitude.GPS, shape=Pathologie, color = oob.score), # Map aesthetics for points
             size = 3, alpha = 1)+
  geom_jitter()+
  scale_colour_gradient2(mid = "yellow", high = "red", na.value = NA, midpoint = 0.45)+
  theme_minimal()

ggplot() +
  ggtitle("Fungal endophyte")+
  geom_sf(data = quebec_sf, fill = "lightblue", color = "black") +
    coord_sf(xlim = c(-80, -70), ylim = c(44.8, 47.5), expand = FALSE)+
    geom_point(data = fmeta_dys_inf, aes(x = Longitude.GPS, y = Latitude.GPS, shape=Pathologie, color = oob.score), # Map aesthetics for points
               size = 3, alpha = 1)+
    geom_jitter()+
    scale_colour_gradient2(mid = "yellow", high = "red", na.value = NA, midpoint = 0.45)+
    theme_minimal()

```

### GTDs exploration of other important pathogens

```{r, results = 'hide',  message=FALSE, warning=FALSE }

library(magrittr)

# Fungal pathogens
table(MBVITS_mt$tax_table$Genus=="Phaeomoniella")
table(MBVITS_mt$tax_table$Genus=="Diplodia")
table(MBVITS_mt$tax_table$Genus=="Phomopsis")
table(MBVITS_mt$tax_table$Genus=="Fusicoccum")
table(MBVITS_mt$tax_table$Genus=="Phaeoacremonium")#present
table(MBVITS_mt$tax_table$Genus=="Cadophora")#present
table(MBVITS_mt$tax_table$Genus=="Cylindrocarpon")
table(MBVITS_mt$tax_table$Genus=="Fusarium")#present
table(MBVITS_mt$tax_table$Genus=="Phaeomoniella")

#if this are more present in infected plant than healthy plants
#Neofusicoccum parvum
#Diplodia mutila
#Phomopsis viticola
#Fusicoccum aesculi
#Phaeoacremonium minimum
#Cadophora melinii
#Phaeomoniella chlamydospora
#Cylindrocarpon
#Fusarium oxysporum
#data
MBV16S_mt$cal_betadiv()
# extract two phyla to show the steps
d1 <- clone(MBVITS_mt)
d1$tax_table <- d1$tax_table[d1$tax_table$Genus == "Phaeoacremonium", ]
d1$tidy_dataset()
d1$cal_betadiv()
d2 <- clone(MBVITS_mt)
d2$tax_table <- d2$tax_table[d2$tax_table$Genus == "Cadophora", ]
d2$tidy_dataset()
d2$cal_betadiv()

d3 <- clone(MBVITS_mt)
d3$tax_table <- d3$tax_table[d3$tax_table$Genus == "Fusarium", ]
d3$tidy_dataset()
d3$cal_betadiv()

# first perform mantel test
t1 <- trans_env$new(dataset = d1, env_cols = 28:36)
t1$data_env <- na.omit(t1$data_env)
t1$cal_mantel(use_measure = "bray", partial_mantel = TRUE)
t2 <- trans_env$new(dataset = d2, env_cols = 8:15)
t2$cal_mantel(use_measure = "bray", partial_mantel = TRUE)
t3 <- trans_env$new(dataset = d3, env_cols = 8:15)
t3$cal_mantel(use_measure = "bray", partial_mantel = TRUE)
# extract a part of the results 
x1 <- data.frame(spec = "Phaeoacremonium", t1$res_mantel) %>% .[, c(1, 3, 6, 8)]
x2 <- data.frame(spec = "Cadophora", t2$res_mantel) %>% .[, c(1, 3, 6, 8)]
x3 <- data.frame(spec = "Fusarium", t3$res_mantel) %>% .[, c(1, 3, 6, 8)]
# rename columns
colnames(x1) <- colnames(x2) <- colnames(x3)<- c("spec", "env", "r", "p.value")
# generate interval data
x1 <- dplyr::mutate(x1, rd = cut(r, breaks = c(-Inf, 0.3, 0.6, Inf), labels = c("< 0.3", "0.3 - 0.6", ">= 0.6")),
                      pd = cut(p.value, breaks = c(-Inf, 0.01, 0.05, Inf), labels = c("< 0.01", "0.01 - 0.05", ">= 0.05")))
x2 <- dplyr::mutate(x2, rd = cut(r, breaks = c(-Inf, 0.3, 0.6, Inf), labels = c("< 0.3", "0.3 - 0.6", ">= 0.6")),
                      pd = cut(p.value, breaks = c(-Inf, 0.01, 0.05, Inf), labels = c("< 0.01", "0.01 - 0.05", ">= 0.05")))
x3 <- dplyr::mutate(x3, rd = cut(r, breaks = c(-Inf, 0.3, 0.6, Inf), labels = c("< 0.3", "0.3 - 0.6", ">= 0.6")),
                    pd = cut(p.value, breaks = c(-Inf, 0.01, 0.05, Inf), labels = c("< 0.01", "0.01 - 0.05", ">= 0.05")))
# cobine two tables
plot_table <- rbind(x1, x2, x3)
# install ggcor following the steps (https://chiliubio.github.io/microeco_tutorial/intro.html#github-packages)
library(ggplot2)

#install.packages("GGally")

library(GGally)

library(ggplot2)

ggplot(plot_table, aes(x = env, y = spec, fill = r)) +
  geom_tile(color = "white") +
  
  # show correlation values
  geom_text(aes(label = round(r, 2)), size = 3) +
  
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1),
    name = "Correlation (r)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(x = "Environmental variables", y = "Taxa")

ggplot(plot_table, aes(x = env, y = spec, fill = r)) +
  geom_tile(color = "white") +
  
  geom_text(
    aes(label = ifelse(p.value < 0.05, round(r, 2), "")),
    size = 3
  ) +
  
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_table$stars <- ifelse(plot_table$p.value < 0.01, "***",
                   ifelse(plot_table$p.value < 0.05, "**", ""))

ggplot(plot_table, aes(x = env, y = spec, fill = r)) +
  geom_tile(color = "white") +
  
  geom_text(aes(label = paste0(round(r, 2), stars)), size = 3) +
  
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Dysbiosis confusion matrix

```{r, results = 'hide',  message=FALSE, warning=FALSE }
#Confusion Matrix
aggregate(CentroidDist_score~Infected,  data = dysbiosis_results_bacteria$dysbiosisEuclideandist,  FUN = mean)

dysbiosis_results_bacteria$dysbiosisOOBRF <- dysbiosis_results_bacteria$dysbiosisOOBRF %>% 
  mutate(OOBPredicted = if_else(oob.score >=0.5, "Infected", "Healthy"))

dysbiosis_results_bacteria$dysbiosisOOBRF <- dysbiosis_results_bacteria$dysbiosisOOBRF %>% 
  mutate(OOBPredicted = if_else(oob.score >=0.5, "Infected", "Healthy"))

#calculate the mean
aggregate(log2Stats~Infected,  data = dysbiosis_results_fungi$dysbiosisCLOUD,  FUN = mean)
dysbiosis_results_fungi$dysbiosisOOBRF <- dysbiosis_results_fungi$dysbiosisOOBRF %>% 
  mutate(OOBPredicted = if_else(oob.score >=0.5, "Infected", "Healthy"))

dysbiosis_results_fungi$dysbiosisOOBRF <- dysbiosis_results_fungi$dysbiosisOOBRF %>% 
  mutate(OOBPredicted = if_else(oob.score >=0.5, "Infected", "Healthy"))
#install.packages("caret")
library(caret)
# 'data' is predicted classes, 'reference' is actual classes
dysbiosis_results_bacteria$dysbiosisOOBRF$OOBPredicted <- as.factor(dysbiosis_results_bacteria$dysbiosisOOBRF$OOBPredicted)

cmb <- confusionMatrix(data = dysbiosis_results_bacteria$dysbiosisOOBRF$OOBPredicted, reference = dysbiosis_results_bacteria$dysbiosisOOBRF$Infected)
print(cmb)

dysbiosis_results_fungi$dysbiosisOOBRF$OOBPredicted <- as.factor(dysbiosis_results_fungi$dysbiosisOOBRF$OOBPredicted)

cmf <- confusionMatrix(data = dysbiosis_results_fungi$dysbiosisOOBRF$OOBPredicted, reference = dysbiosis_results_fungi$dysbiosisOOBRF$Infected)
print(cmf)
```

### Ecological assembly processes

```{r ecoa,  message=FALSE, warning=FALSE}
#### ecological processes dominating the community assembly 
var.data = data.frame(MBV16S_mt$sample_table)
# generate trans_nullmodel object
# as an example, we only use high abundance OTU with mean relative abundance > 0.0005
t1 <- trans_nullmodel$new(MBV16S_mt, filter_thres = 0.0005, add_data = var.data)

# RCbray only
t1$cal_rcbray(runs = 1000)
rc <- t1$res_rcbray

process <- ifelse(rc > 0.95, "Dispersal limitation",
                  ifelse(rc < -0.95, "Homogenizing dispersal",
                         "Drift / Undominated"))


library(reshape2)

rc_mat <- t1$res_rcbray

# convert to long format
rc_long <- melt(rc_mat, varnames = c("Sample1", "Sample2"), value.name = "RCbray")

# remove self-comparisons
rc_long <- rc_long[rc_long$Sample1 != rc_long$Sample2, ]

meta <- MBV16S_mt$sample_table

meta$OOBPredicted <-dysbiosis_results_bacteria$dysbiosisOOBRF$oob.score 
#meta$OOBPredicted <-dysbiosis_results_fungi$dysbiosisOOBRF$OOBPredicted

rc_long$Tissue1 <- meta[rc_long$Sample1, "Tissu"]
rc_long$Tissue2 <- meta[rc_long$Sample2, "Tissu"]

rc_long$Health1 <- meta[rc_long$Sample1, "Infected"]   
rc_long$Health2 <- meta[rc_long$Sample2, "Infected"]

rc_long$PredHealth1 <- meta[rc_long$Sample1, "OOBPredicted"]   
rc_long$PredHealth2 <- meta[rc_long$Sample2, "OOBPredicted"]
#same tissue
rc_long_tissue <- rc_long[rc_long$Tissue1 == rc_long$Tissue2, ]

rc_long_group <- rc_long[
  rc_long$Tissue1 == rc_long$Tissue2 &
    rc_long$Health1 == rc_long$Health2 &
    rc_long$PredHealth1 == rc_long$PredHealth2, ]

#classify process
rc_long_group$Process <- ifelse(rc_long_group$RCbray > 0.95, "Dispersal limitation",
                                ifelse(rc_long_group$RCbray < -0.95, "Homogenizing dispersal",
                                       "Drift / Undominated"))

rc_long_group$Tissue <- rc_long_group$Tissue1
rc_long_group$Health <- rc_long_group$Health1
rc_long_group$PredHealth <- rc_long_group$PredHealth1


summary_df <- rc_long_group %>%
  group_by(Tissue, Health,  Process) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tissue, Health) %>%
  mutate(Percentage = 100 * n / sum(n))
DT::datatable(summary_df)

summary_df_pred <- rc_long_group %>%
  group_by(Tissue, PredHealth, Process) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tissue, PredHealth) %>%
  mutate(Percentage = 100 * n / sum(n))
DT::datatable(summary_df_pred)


#fungi
t2 <- trans_nullmodel$new(MBVITS_mt, filter_thres = 0.0005, add_data = var.data)

# RCbray only
t2$cal_rcbray(runs = 1000)
rcf <- t2$res_rcbray

processf <- ifelse(rcf > 0.95, "Dispersal limitation",
                  ifelse(rc < -0.95, "Homogenizing dispersal",
                         "Drift / Undominated"))

rc_matf <- t2$res_rcbray

# convert to long format
rc_longf <- melt(rc_matf, varnames = c("Sample1", "Sample2"), value.name = "RCbray")

metaf <- MBVITS_mt$sample_table
metaf$OOBPredicted <-dysbiosis_results_fungi$dysbiosisOOBRF$oob.score[1:186]

rc_longf$Tissue1 <- metaf[rc_longf$Sample1, "Tissu"]
rc_longf$Tissue2 <- metaf[rc_longf$Sample2, "Tissu"]

rc_longf$Health1 <- metaf[rc_longf$Sample1, "Infected"]   
rc_longf$Health2 <- metaf[rc_longf$Sample2, "Infected"]

rc_longf$PredHealth1 <- metaf[rc_longf$Sample1, "OOBPredicted"]  
rc_longf$PredHealth2 <- metaf[rc_longf$Sample2, "OOBPredicted"]
#same tissue
rc_longf_tissue <- rc_longf[rc_longf$Tissue1 == rc_longf$Tissue2, ]

rc_longf_group <- rc_longf[
  rc_longf$Tissue1 == rc_longf$Tissue2 &
    rc_longf$Health1 == rc_longf$Health2 &
    rc_longf$PredHealth1 == rc_longf$PredHealth2, ]

#classify process
rc_longf_group$Process <- ifelse(rc_longf_group$RCbray > 0.95, "Dispersal limitation",
                                ifelse(rc_longf_group$RCbray < -0.95, "Homogenizing dispersal",
                                       "Drift / Undominated"))

rc_longf_group$Tissue <- rc_longf_group$Tissue1
rc_longf_group$Health <- rc_longf_group$Health1
rc_longf_group$PredHealth <- rc_longf_group$PredHealth1


summary_dff <- rc_longf_group %>%
  group_by(Tissue, Health,  Process) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tissue, Health) %>%
  mutate(Percentage = 100 * n / sum(n))
DT::datatable(summary_dff)

summary_df_predf <- rc_longf_group %>%
  group_by(Tissue, PredHealth, Process) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tissue, PredHealth) %>%
  mutate(Percentage = 100 * n / sum(n))
DT::datatable(summary_df_predf)


```

### Structural Equation Modelling

```{r sem,  message=FALSE, warning=FALSE, eval=FALSE}
#SEM
library(piecewiseSEM)

#data
library(dplyr)
library(tidyr)

alpha_bact <- MBV16S_alpha$data_alpha %>%
  dplyr::filter(Measure %in% c("Observed", "Shannon_hill", "Simpson_hill")) %>%
  dplyr::select(Sample, Measure, Value) %>%
  tidyr::pivot_wider(
    names_from = Measure,
    values_from = Value
  )

alpha_fung <- MBVITS_alpha$data_alpha %>%
  dplyr::filter(Measure %in% c("Observed", "Shannon_hill", "Simpson_hill")) %>%
  dplyr::select(Sample, Measure, Value) %>%
  tidyr::pivot_wider(
    names_from = Measure,
    values_from = Value
  )
colnames(ord_scores_bact)[3] <- "Sample"
colnames(ord_scores_fung)[3] <- "Sample"
alpha_bact <- alpha_bact %>% rename_with(~ paste0(.x, "_bact"), -Sample)
alpha_fung <- alpha_fung %>% rename_with(~ paste0(.x, "_fung"), -Sample)
dim(ord_scores_bact)

dfbact <- ord_scores_bact %>%
  left_join(alpha_bact, by = "Sample")
dffung <-  ord_scores_fung %>%
  left_join(alpha_fung, by = "Sample")

colnames(dffung)[41]<-"dysbiosis"


# Clean and select relevant variables
df_clean <- dfbact %>%
#  df_clean <- dffung %>%
  dplyr::select(
    Sample,
    
    # --- Environment (keep a few representative variables)
    Mean.Temp,
    Mean.humidity,
    Mean.Wind.speed,
    Latitude.GPS,
    Longitude.GPS,
    
    # --- Pathogens
    Eutypa.lata,
    Diplodia.seriata,
    
    # --- Infection status
    Infected,
    
    # --- Dysbiosis
    dysbiosis,
    
    # --- Diversity bacteria
    Observed_bact,
    Shannon_hill_bact,
    Simpson_hill_bact,
    # --- Diversity fungi
    #Observed_fung,
    #Shannon_hill_fung,
    #Simpson_hill_fung,
    
    # --- Community structure
    PCo1,
    PCo2
  )


#clean coordinates
df_clean <- df_clean %>%
  dplyr::select(-Latitude.GPS, -Longitude.GPS)
str(df_clean)

#combine pathogens into one latent variable
df_clean$Eutypa.lata <- as.integer(df_clean$Eutypa.lata)
df_clean$Pathogen_load = df_clean$Eutypa.lata + df_clean$Diplodia.seriata

df_clean <- df_clean %>%
  drop_na()


#check collinearity
df_num <- df_clean %>% dplyr::select(where(is.numeric))

cor_mat <- cor(df_num)

# inspect strong correlations
which(abs(cor_mat) > 0.8, arr.ind = TRUE)

#select variables
#Bacteria
df_clean.s <- df_clean %>%
  dplyr::select(Mean.Temp, Mean.humidity, Mean.Wind.speed, dysbiosis, Shannon_hill_bact, PCo1, PCo2, Pathogen_load)
#FUNGI
df_clean.s <- df_clean %>%
  dplyr::select(Mean.Temp, Mean.humidity, Mean.Wind.speed, dysbiosis, Shannon_hill_fung, PCo1, PCo2, Pathogen_load)


df_scaled <- df_clean %>%
  mutate(across(where(is.numeric), ~ as.numeric(scale(.))))


#build SEM
library(piecewiseSEM)

m1 <- lm(dysbiosis ~ Mean.Temp + Mean.humidity + Mean.Wind.speed, data = df_scaled)

m2 <- lm(Shannon_hill_bact ~ dysbiosis + Mean.Temp, data = df_scaled)
m2 <- lm(Shannon_hill_fung ~ dysbiosis + Mean.Temp, data = df_scaled)

m3 <- lm(Pathogen_load ~ dysbiosis + Mean.Temp, data = df_scaled)
m4 <- lm(PCo1 ~ Mean.Temp + Mean.humidity, data = df_scaled)


sem_model <- psem(m1, m2, m3, m4)

summary(sem_model)

library(lavaan)
#install.packages("lavaan")
#install.packages("semPlot")
library(semPlot)

#Define SEM structure
model <- '
  # Dysbiosis drivers
  dysbiosis ~ Mean.Temp + Mean.humidity + Mean.Wind.speed

  # Microbiome response
  Shannon_hill_bact ~ dysbiosis + Mean.Temp

  # Pathogens
  Pathogen_load ~ dysbiosis + Mean.Temp
  
  # Community structure
  PCo1 ~ Mean.Temp + Mean.humidity
'

fit <- sem(model, data = df_scaled)

pe <- parameterEstimates(fit, standardized = TRUE)
paths <- pe %>%
  filter(op == "~") %>%
  dplyr::select(lhs, rhs, est, pvalue)


#build node layout
nodes <- data.frame(
  name = c("Mean.Temp", "Mean.humidity", "Mean.Wind.speed",
           "dysbiosis",
           "Shannon_hill_bact",
           "Pathogen_load",
           "PCo1"),
  
  x = c(1, 1, 1,
        2,
        3,
        3,
        2),
  
  y = c(3, 2, 1,
        2,
        3,
        1,
        1.5)
)


#build edges (with significance styling)

edges <- paths %>%
  mutate(
    sig = ifelse(pvalue < 0.05, "sig", "ns")
  )

# Complete Graph Object
#install.packages("tidygraph")
library(tidygraph)
#install.packages("ggraph")
library(ggraph)
param_graph <- tidygraph::tbl_graph(nodes, edges)

fit <- sem(model, data = df_scaled)

#ggplot2 SEM models with tidygraph and ggraph 
#https://drsimonj.svbtle.com/ggsem-plot-sem-models-with-ggplot2
# Plot a fitted lavaan object
ggsem <- function(fit, layout = "circle") {
  
  # Extract standardized parameters
  params <- lavaan::standardizedSolution(fit)
  
  # Edge properties
  param_edges <- params %>% 
    filter(op %in% c("=~", "~", "~~"), lhs != rhs, pvalue < .10) %>%
    transmute(to = lhs,
              from = rhs,
              val = est.std,
              type = dplyr::case_when(
                op == "=~" ~ "loading",
                op == "~"  ~ "regression",
                op == "~~" ~ "correlation",
                TRUE ~ NA_character_))
  
  # Identify latent variables for nodes
  latent_nodes <- param_edges %>% 
    filter(type == "loading") %>% 
    distinct(to) %>% 
    transmute(metric = to, latent = TRUE)
  
  # Node properties
  param_nodes <- params %>% 
    filter(lhs == rhs) %>% 
    transmute(metric = lhs, e = est.std) %>% 
    left_join(latent_nodes) %>% 
    mutate(latent = if_else(is.na(latent), FALSE, latent))
  
  # Complete Graph Object
  param_graph <- tidygraph::tbl_graph(param_nodes, param_edges)
  
  # Plot
  ggraph(param_graph, layout = layout) +
    # Latent factor Nodes
    geom_node_point(aes(alpha = as.numeric(latent)),
                    shape = 16, size = 14) +
    geom_node_point(aes(alpha = as.numeric(latent)),
                    shape = 16, size = 12, color = "white") +
    # Observed Nodes
    geom_node_point(aes(alpha = as.numeric(!latent)),
                    shape = 15, size = 14) +
    geom_node_point(aes(alpha = as.numeric(!latent)),
                    shape = 15, size = 12, color = "white") +
    # Regression Paths (and text)
    geom_edge_link(aes(color = val, label = round(val, 2),
                       alpha = as.numeric(type == "regression")),
                   linetype = 1, angle_calc = "along", vjust = -.5,
                   arrow = arrow(20, unit(.3, "cm"), type = "closed")) +
    # Factor Loadings (no text)
    geom_edge_link(aes(color = val, alpha = as.numeric(type == "loading")),
                   linetype = 3, angle_calc = "along",
                   arrow = arrow(20, unit(.3, "cm"), ends = "first", type = "closed")) +
    # Correlation Paths (no text)
    geom_edge_link(aes(color = val, alpha = as.numeric(type == "correlation")),
                   linetype = 2, angle_calc = "along",
                   arrow = arrow(20, unit(.3, "cm"), type = "closed", ends = "both")) +
    # Node names
    geom_node_text(aes(label = metric),
                   nudge_y = .25, hjust = "inward") +
    # Node residual error
    geom_node_text(aes(label = sprintf("%.2f", e)),
                   nudge_y = -.1, size = 14) +
    # Scales and themes
    scale_alpha(guide = FALSE, range = c(0, 1)) +
    scale_edge_alpha(guide = FALSE, range = c(0, 1)) +
    scale_edge_colour_gradient2(guide = FALSE, low = "red", mid = "darkgray", high = "green") +
    scale_edge_linetype(guide = FALSE) +
    scale_size(guide = FALSE) +
    theme_graph()
}

#modified by me
ggsem <- function(fit, layout = "tree") {
  
  # Extract standardized parameters
  params <- lavaan::standardizedSolution(fit)
  
  # Edge properties
  param_edges <- params %>% 
    filter(op %in% c("=~", "~", "~~"), lhs != rhs, pvalue < .10) %>%
    transmute(to = lhs,
              from = rhs,
              val = est.std,
              type = dplyr::case_when(
                op == "=~" ~ "loading",
                op == "~"  ~ "regression",
                op == "~~" ~ "correlation",
                TRUE ~ NA_character_))
  
  # Identify latent variables for nodes
  latent_nodes <- param_edges %>% 
    filter(type == "loading") %>% 
    distinct(to) %>% 
    transmute(metric = to, latent = TRUE)
  
  # Node properties
  param_nodes <- params %>% 
    filter(lhs == rhs) %>% 
    transmute(metric = lhs, e = est.std) %>% 
    left_join(latent_nodes) %>% 
    mutate(latent = if_else(is.na(latent), FALSE, latent))
  
  # Complete Graph Object
  param_graph <- tidygraph::tbl_graph(param_nodes, param_edges)
  
  # Plot
  ggraph(param_graph, layout = layout) +
    # Latent factor Nodes
    geom_node_point(aes(alpha = as.numeric(latent)),
                    shape = 16, size = 8) +
    geom_node_point(aes(alpha = as.numeric(latent)),
                    shape = 16, size = 8, color = "white") +
    # Observed Nodes
    geom_node_point(aes(alpha = as.numeric(!latent)),
                    shape = 15, size = 8) +
    geom_node_point(aes(alpha = as.numeric(!latent)),
                    shape = 15, size = 8, color = "white") +
    # Regression paths
    geom_edge_arc(
      aes(color = val, label = round(val, 2),
          alpha = as.numeric(type == "regression")),
      strength = 0.4,
      arrow = arrow(20, unit(.3, "cm"), type = "closed"),
      label_size = 4
    ) +
    # Loadings
    geom_edge_arc(
      aes(color = val,
          alpha = as.numeric(type == "loading")),
      strength = 0.3,
      linetype = 3,
      arrow = arrow(20, unit(.3, "cm"), ends = "first", type = "closed")
    ) +
    # Correlations
    geom_edge_arc(
      aes(color = val,
          alpha = as.numeric(type == "correlation")),
      strength = 0.6,
      linetype = 2,
      arrow = arrow(20, unit(.3, "cm"), ends = "both", type = "closed")
    ) +
    # Node names
    geom_node_text(aes(label = metric),
                   nudge_y = .25, hjust = "inward") +
    # Node residual error
    geom_node_text(aes(label = sprintf("%.2f", e)),
                   nudge_y = -.1, size = 10) +
    # Scales and themes
    scale_alpha(guide = FALSE, range = c(0, 1)) +
    scale_edge_alpha(guide = FALSE, range = c(0, 1)) +
    scale_edge_colour_gradient2(guide = FALSE, low = "red", mid = "darkgray", high = "green") +
    scale_edge_linetype(guide = FALSE) +
    scale_size(guide = FALSE) +
    theme_graph()
}

#install.packages('ggsem')
#plot SEM
ggsem(fit)

ggsem::ggsem(fit)


# Diagram
library(DiagrammeR)

grViz("
digraph SEM {
  graph [layout = dot, rankdir = LR, overlap = false, splines = true]

  # Node definitions with styles
  node [shape = rectangle, style = filled, fillcolor = White, fontname = Helvetica, penwidth = 2]
  
  # Predictors (Left)
  subgraph cluster_0 {
    graph [peripheries = 0]
    'Mean Temperature'; 'Mean humidity'; 'Mean Wind speed';
  }

  # Mediator (Middle)
  'Dysbiosis';

  # Responses (Right)
  'Shannon Hill'; 'Pathogen Occurence'; 'PCo1';

  # Edges from model string
  'Mean Temperature' -> 'Dysbiosis'
  'Mean humidity' -> 'Dysbiosis'
  'Pathogen Occurence' -> 'Dysbiosis'
  'PCo1' -> 'Dysbiosis'
  
  'Dysbiosis' -> 'Shannon Hill'
  'Mean Temperature' -> 'Shannon Hill'
  
  'Dysbiosis' -> 'Pathogen Occurence'
  'Mean Temperature' -> 'Pathogen Occurence'
  
  'Mean Temperature' -> 'PCo1'
  'Mean humidity' -> 'PCo1'
  'Mean Wind speed' -> 'PCo1'
}
")

```
